---
title: "NFL Fatigue Analysis in Defensive Linemen"
author: "Sarah Deussing & Imogen Meers"
date: "2024-11-11"
output:
  pdf_document: default
  html_document: default
---

## Motivation
Our analysis will study fatigue in defensive linemen using their speed, acceleration, distance traveled, and orientation. We will be calculating a player load and analyze the defensive player's effectiveness. To do so, we will combine this movement data with play outcome data - if the player sacked the quarterback, had a solo tackle, etc.

The outcome of our analysis will be to predict likelihood of "mistakes" and positive plays, based on player fatigue and other game-time variables and to identify if specific players are more fatigue-resistant.

There has been published research on long-term fatigue and "burnout" in NFL players over the course of their careers, which focuses on long-term health and physiological effects. There has also been research on game day fatigue in European Football to determine limitations on performance based on fatigue, in order to create suitable, enhancing interventions (Alghannam). However, there is no player specific or NFL game day based fatigue analysis, which is what will set our project apart.


## Research
We wanted to find a measure of fatigue that was derived from tracking data. From an study on _Monitoring Fatigue During Intermittent Exercise With Accelerometer-Derived Metrics_ and Catapult's own description of Player Load, we found this idea of "Dynamic Load", that is a sum of triaxial acceleration across a set time period. We do not have acceleration in three plains but we do have an acceleration in the direction of play so we decided to derive a "Load per Play" value that will be acceleration summed per player per play.

As an extension to this, we will also derive a "Fatigue Index" based on the research by Beato et al., that will be a value of acceleration/speed for each value in time, i.


## Problem framing
We will be creating two derived variables:

- Load by player by play
- Fatigue index by player by play

Perform a time-series analysis over the length of a game/period to see trend and seasonality of fatigue and the change in fatigue at specific events, e.g. "mistakes".

Define mistake as penalties that will be grouped based on type:
- Formation
  - Illegal Shift, Illegal Formation, Defensive Offside

- Tackle
  - Defensive Holding, Horse Collar Tackle, Low Block, Illegal Use of Hands, Roughing the Passer

Use player demographic data, time, tracking and load data to predict probability of "mistake" at a given time.

Use player demographic data, time, tracking and load data to predict positive defensive outcomes such as sacks, pressure and fumbles.

Identify fatigue-resistant players and cluster to see if they have similar demographic characteristics.

## Data overview
The dataset is a part of the NFL Big Data Bowl 2025, and it is taken from:
https://www.kaggle.com/competitions/nfl-big-data-bowl-2025/data 

We are going to use four main parts of the dataset: player_play, players, plays, and the tracking data for each week.

The player_play data contains information about the action of the play, including its outcome or any penalties. Each row corresponds to one player's individual actions for each play within a game.

The players data contains information about each player, including position, height, weight, and college. We will use this data to filter to only defensive linemen within all data sets.

The plays data contains information about the quarter, time, and contents of each play.

The tracking data has the x-coordinate, y-coordinate, direction, orientation, speed, and acceleration for each player involved in each play in a game. We will use this movement data to calculate player fatigue. This tracking data is available for all 9 weeks of the season.

The datasets can be joined on: gameId, playId, and nflId.

## Contribution
Imogen Meers & Sarah Deussing

## Bibliography
Alghannam, Abdullah F. “Metabolic limitations of performance and fatigue in football.” Asian journal of sports medicine vol. 3,2 (2012): 65-73. doi:10.5812/asjsm.34699

Beato, Marco et al. “Monitoring Fatigue During Intermittent Exercise With Accelerometer-Derived Metrics.” Frontiers in physiology vol. 10 780. 26 Jun. 2019, doi:10.3389/fphys.2019.00780

Finster-Rowen, Alex. "What is Player Load?" Catapult Sports. Catapult Sports, n.d. Web.


## Implementation
```{r}
library(dplyr)
library(ggplot2)
library(lubridate)
library(tidyr)
library(gridExtra)
library(dendextend)
```

#### Data
Play Data for Each Player
```{r read data}
play <- read.csv('player_play.csv')
player_data <- read.csv('players.csv')
```

Filter to only defensive linemen.
- DE (Defensive End)
- DT (Defensive Tackle)
- NT (Nose Tackle)

```{r make def line}
def_line_players <- player_data[player_data$position == 'DE' | player_data$position == 'DT' |
                          player_data$position == 'NT',]

def_line <- play %>%
  left_join(def_line_players[, c("nflId", "position")], by = "nflId")

def_line <- def_line %>%
  select(-c('hadRushAttempt', 'rushingYards', 'hadDropback', 'passingYards', 'sackYardsAsOffense',
            'hadPassReception', 'receivingYards', 'wasTargettedReceiver', 'yardageGainedAfterTheCatch',
            'fumbles', 'fumbleLost', 'fumbleOutOfBounds', 'wasInitialPassRusher', 'wasRunningRoute',
            'routeRan', 'blockedPlayerNFLId1', 'blockedPlayerNFLId2', 'blockedPlayerNFLId3'))
```

Variables for defensive linemen: 
 - assistedTackle: required an assist to make a tackle
 - forcedFumbleAsDefense: forced a fumble by the opposing team
 - halfSackYardsAsDefense: yards conceded by the offense because of a half-sack
 - passDefensed: a passing play was stopped by the player
 - quarterbackHit: player recorded a QB hit
 - sackYardsAsDefense: yards conceded by the offense because of a sack by the player
 - safetyAsDefense: player forced a safety on this play
 - soloTackle: player recorded a solo tackle on this play
 - tackleAssist: player was awarded an assisted tackle
 - tackleForALoss: player recorded a tackle behind the line of scrimmage
 - tackleForALossYardage: yards conceded by the offense because of a tackle
 - hadInterception: player intercepted a pass
 - interceptionYards: yards returned by the player on an intercepted pass
 - fumbleRecoveries: number of fumbles recovered by the player
 - fumbleRecoveryYards: yards returned by the player on a fumble recovery
 - penaltyNames
 - causedPressure: player pressured the QB
 - timeToPressureAsPassRusher: time elapsed from snap to player reaching a pressure pro. >= to 0.75
 - inMotionAtBallSnap: player was in motion at snap
 - shiftSinceLineset: player shifted since the lineset
 - pressureAllowedAsBlocker: any pass rushers that the blocker had a true matchup against recorded a pressure
 - timeToPressureAllowedAsBlocker: time from snap to pass rusher who blocker had true matchup against achieving a pressure prob. > 0.75
 - pff_defensiveCoverageAssignment: defensive coverage assignment given to the player
 
```{r}
summary(def_line)
```

Tracking Data
```{r}
week1 <- read.csv('tracking_week_1.csv')
week2 <- read.csv('tracking_week_2.csv')
week3 <- read.csv('tracking_week_3.csv')
weeks <- rbind(week1, week2, week3)
#save(weeks, file = "weeks123.RData")
```

The following variables from the tracking data can be used to calculate fatigue:
 - s (speed), a (acceleration), dis (distance traveled), dir (direction), o (orientation) 
 
Join Data
```{r}
data <- inner_join(def_line, weeks, by = c("gameId", "playId", "nflId"))
#save(data, file = "joinedData.RData")
```

```{r}
# load("joinedData.RData")
colnames(data)
head(data)
```

Two variables of importance are a player's load and fatigue index, which is equal to acceleration/speed. Let's visualize these for one player in one play. We want to look at movement only during/after the ball is snapped. We will filter the frameId to be within a small range of the snap and initial acceleration.

Load Per Play
```{r}
ran_play <- data %>%
  filter(gameId == 2022090800) %>%
  filter(playId == 56) %>%
  filter(frameType != "BEFORE_SNAP") %>%
  filter(!is.na(position)) %>%
  filter(frameId <= 155) %>%
  mutate(fatigueIndex = a/s)
ggplot(ran_play, aes(x=frameId, y = fatigueIndex, color = position)) + geom_point() + geom_smooth()
```

Acceleration Per Play
```{r}
ran_play2 <- data %>%
  filter(gameId == 2022090800) %>%
  filter(playId == 56) %>%
  filter(frameType != "BEFORE_SNAP") %>%
  filter(!is.na(position)) %>%
  filter(frameId <= 160)
ggplot(ran_play2, aes(x=frameId, y = a, color = position)) + geom_point() + geom_smooth()
```

#### Derived Variables

Our derived variables will be calculated on a per-play level.

Player Load = SUM(Acceleration)
Max Acceleration = MAX(Acceleration)
```{r calculate load}
data <- data %>%
  filter(frameType != 'BEFORE_SNAP') %>%
  group_by(gameId, playId, nflId) %>%
  mutate(load = sum(a, na.rm = TRUE),
         max_acc = max(a)) %>% ungroup()

head(data)
```

Cumulative Load = CUMSUM(load)
```{r calculate cumulative load}
per_play <- data %>%
  select(gameId, playId, nflId, load, teamAbbr, position, max_acc) %>% distinct() %>%
  arrange(gameId, nflId, playId)
#head(per_play)

per_play <- per_play %>%
  group_by(gameId, nflId) %>%
  mutate(cum_load = cumsum(load))
head(per_play)
```

#### Exploratory Analysis/Visualizations

Load Per Player
```{r load per player plot}
game <- data[data$gameId == 2022090800,]

# select 10 defensive linemen at random
set.seed(456)
ran_nflIds <- sample(game$nflId, 10)

for (id in ran_nflIds) {
  player_data <- game %>% filter(nflId == id) %>% filter(frameId <= 160)
  load_plot <- ggplot(player_data, aes(x = playId, y = load)) +
    geom_point() + 
    geom_smooth(method = "loess", color = "blue") + 
    labs(x = "Play ID", y = "Load", title = paste0("Player Load in Game - NFL ID: ", id)) +
    theme_minimal()

  print(load_plot)
}
```

We hypothesized that total load would decrease later in the game because a player would be more fatigues. For most of the linemen, we don't see a decline in load throughout the game. These players either remain at a almost steady load value or increase slightly towards the end of the game. This trend could be the result of a break at halftime or a higher-intensity end to a game. 

Let's instead look at cumulative load over a game. 
```{r}
game <- per_play[per_play$gameId == 2022090800,]

# select 10 defensive linemen at random
set.seed(456)
ran_nflIds <- sample(game$nflId, 10)

for (id in ran_nflIds) {
  player_data <- game %>% filter(nflId == id)
  c_load_plot <- ggplot(player_data, aes(x = playId, y = cum_load)) +
    geom_smooth(method = "loess", color = "blue") + 
    labs(x = "Play ID", y = "Cumulative Load", title = paste0("Player Cum. Load in Game - NFL ID: ", id)) +
    theme_minimal()

  print(c_load_plot)
}
```

In addition to calculating total load using the sum of accelerations, we can also look at the change in the fatigue analysis of a player. This measure will incorporate both player speed and acceleration during the play.

Fatigue Index
```{r}
for (id in ran_nflIds) {
  player_data <- game %>% filter(nflId == id) %>% filter(frameType != "BEFORE_SNAP") %>% mutate(fatigueIndex = a/s)
  load_plot <- ggplot(player_data, aes(x = playId, y = fatigueIndex)) +
    geom_point() + 
    geom_smooth(method = "loess", color = "blue") + 
    labs(x = "Play ID", y = "Load", title = paste0("Player Fatigue Index in Game - NFL ID: ", id)) +
    theme_minimal()

  print(load_plot)
}
```

Player height and weight may also have a relationship with acceleration. Let's visualize this relationship.

Effect of Demographics on Median Acceleration
```{r}
max_acc <- data %>% group_by(nflId) %>% mutate(max_acc = mean(a)) %>% select(nflId, max_acc) %>% distinct()
demographic <- max_acc %>% inner_join(def_line_players)
demographic$birthDate <- as.Date(demographic$birthDate)


# Function to convert feet-inches to centimeters 
convert_to_cm <- function(height)
  { height_parts <- strsplit(height, "-")[[1]] 
  feet <- as.numeric(height_parts[1]) 
  inches <- as.numeric(height_parts[2]) 
  total_inches <- feet * 12 + inches 
  total_cm <- total_inches * 2.54 
  return(total_cm) } # Apply the conversion to the height column 

demographic$height <- sapply(demographic$height, convert_to_cm)

demographic <- demographic %>% na.omit() %>% mutate(age = as.integer(round(as.Date(now()) - birthDate)/365))

# Scatter plot for age vs x
plot_age <- ggplot(demographic, aes(x = age, y = max_acc)) +
  geom_point() +
  labs(title = "Age vs Acceleration", x = "Age", y = "Avg Acceleration") +
  theme_minimal() +geom_smooth()

# Scatter plot for height vs x
plot_height <- ggplot(demographic, aes(x = height, y = max_acc)) +
  geom_point() +
  labs(title = "Height vs Acceleration", x = "Height", y = "Avg Acceleration") +
  theme_minimal() +geom_smooth()

# Scatter plot for weight vs x
plot_weight <- ggplot(demographic, aes(x = weight, y = max_acc)) +
  geom_point() +
  labs(title = "Weight vs Acceleration", x = "Weight", y = "Avg Acceleration") +
  theme_minimal() +geom_smooth()

grid.arrange(plot_age, plot_height, plot_weight, nrow = 1)

```

As one of the points of our analysis is to look for key factors that affect fatigue-resistant, we plotted the relationship between age, height and weight and median acceleration. As acceleration will be included in our derivation for "fatigue", we thought it would be interesting to see these relationships.

We would have thought as players get older and larger that acceleration decreases. However this was not the case, with only weight showing a marginally negative correlation. This probably means that other factors such as match-up and time played/time since last TO are more important factors.

With our fatigue analysis, we will also be looking at the outcome of the play. We have divided penalties into two main categories: tackle penalties (which happen during the play) and formation penalties (which happen before the play).
```{r penalties}
tackle_pen <- c("Horse Collar Tackle", "Defensive Holding", "Low Block", "Illegal Use of Hands", "Roughing the Passer")
formation_pen <-  c("Illegal Shift", "Illegal Formation", "Defensive Offside")
```


Penalties Throughout a Game
```{r Tackles over Time}
play_details <- read.csv("plays.csv") %>% select(gameId, playId, quarter, gameClock)
head(play_details)

# penalties <- play %>% select(gameId, playId, nflId, penaltyNames) %>% full_join(play_details) %>% mutate(penaltyType = ifelse(penaltyNames %in% tackle_pen, "Tackle", ifelse(penaltyNames %in% formation_pen, "Formation", NA)), totalClock = (quarter-1) *  ms("15:00") +(ms("15:00")- ms(gameClock))) 


penalties <- play %>% select(gameId, playId, nflId, penaltyNames) %>% full_join(play_details) %>% mutate(penaltyType = ifelse(penaltyNames %in% tackle_pen, "Tackle", ifelse(penaltyNames %in% formation_pen, "Formation", NA)))
                                                                                                         
plot_data <- penalties %>% select(quarter, penaltyType) %>% na.omit() %>% group_by(quarter, penaltyType) %>% summarise(count = n()) %>% ungroup() 

stacked_bar <- ggplot(plot_data, aes(x = as.factor(quarter), y = count, fill = penaltyType)) +
  geom_bar(stat = "identity") +
  labs(title = "Penalties by Category and Quarter", x = "Quarter", y = "Count", fill = "Penalty Type") +
  theme_minimal()

stacked_bar

# plot_data <- plot_data %>% na.omit()
# plot_data <- plot_data %>% pivot_wider(names_from = penaltyType, values_from = count, values_fill = list(count = 0))
# 
# plot_data <- penalties %>% select(gameId, playId, penaltyType) %>% distinct() %>% arrange(gameId)
# unique(plot_data$playId)
# 
# ggplot(data = plot_data) + geom_density(aes(x = totalClock, y = count, color = penaltyType))
# tail(penalties)
# unique(penalties$totalClock)
# ms("15:00")- ms(penalties$gameClock)

```

This doesn't show exactly what we thought. There is an increase in both types of penalties from the first to the second quarter, and an increase in formation penalties from the third to fourth quarter. We would expect some type of reset after half time, so we will also look at fatigue by quarter. This is just a plot for week 1 data so it's not totally representative of the season or the "NFL" in general.

#### Clustering Analysis
We are going to cluster players based on their:
  - cumulative load
  - fatigueIndex
  - number of penalties
  - height
  - weight
  - age
  - max. acceleration

```{r}
# for each gameId, playId, nflId group: if any of the penalties match a value in tackle or formation penalties, set this value to 1
total_penalties <- function(penaltyNames, tackle_pen, formation_pen) {
  tacklePenalty = ifelse(any(penaltyNames %in% tackle_pen), 1, 0)
  formationPenalty = ifelse(any(penaltyNames %in% formation_pen), 1, 0)
  return(c(tacklePenalty, formationPenalty))
}

grouped <- data %>%
  group_by(gameId, playId, nflId) %>%
  mutate(
    penalty_values = list(total_penalties(penaltyNames, tackle_pen, formation_pen)),
    tackle_pen = penalty_values[[1]][1], 
    formation_pen = penalty_values[[1]][2],
    avg_fatigueIndex = sum(mean(a)/mean(s))) %>%
  select(gameId, playId, nflId, displayName, frameId, load, max_acc, penalty_values, tackle_pen, formation_pen, avg_fatigueIndex)
```

Some values of average fatigue index are Infinite because the player's logged speed is NaN. Let's impute 0 for those values.
```{r}
grouped$avg_fatigueIndex[is.infinite(grouped$avg_fatigueIndex)] <- 0
```


```{r, eval=FALSE}
# check
grouped %>%
  filter(tackle_pen == 1 | formation_pen == 1)
```

We need to add cumulative load and fatigueIndex to this dataset, making it on a per-play level.
```{r}
test <- grouped %>%
  select(gameId, playId, nflId, load, displayName, max_acc, tackle_pen, formation_pen, avg_fatigueIndex) %>%
  distinct() %>%
  arrange(gameId, nflId, playId)

test <- test %>%
  group_by(gameId, nflId) %>%
  mutate(cum_load = cumsum(load))

head(test)
```

We also need to add player height, weight, and age.
```{r}
test <- test %>%
  left_join(player_data[, c("nflId", "height", "weight", "birthDate")], by = "nflId")
```

Now that we have a dataset with the necessary values for our clustering, we need to clean the height and birthDate columns. We will change height to an numeric inches value and birthDate to a numerical age.
```{r}
# height conversion
class(test$height)
convert_to_inches <- function(height_str) {
  parts <- strsplit(height_str, "-")[[1]]
  feet <- as.numeric(parts[1])
  inches <- as.numeric(parts[2])
  total_inches <- (feet * 12) + inches
  return(total_inches)
}
test$height <- sapply(test$height, convert_to_inches)

# birthDate conversion
class(test$birthDate)
convert_to_age <- function(birthdate) {
  birthdate <- as.Date(birthdate, format = "%Y-%m-%d")
  date <- Sys.Date()
  age <- as.numeric(difftime(Sys.Date(), birthdate, units = "weeks")) / 52.25
  return(floor(age))
}
test$age <- convert_to_age(test$birthDate)
```

We want to analyze only players that occur many times in the dataset. We will eliminate players that only have a few total entries in the dataset, as these are not the starting players.
```{r}
players_top <- test %>% ungroup() %>% select(displayName) %>%
  group_by(displayName) %>% mutate(count = n()) %>% distinct()
nrow(players_top)
players_top <- players_top %>% filter(count >= 100)
nrow(players_top)
```

By selecting players with over 100 instances in the dataset, we have cut our dataset by over 50%. Selecting only these players will eliminate outliers - instances in which a player had only a few movements in a game.

Select Columns to Cluster.
```{r select cols}
clustering_columns <- c("max_acc", "avg_fatigueIndex", "cum_load", "height", "weight", "age", "load")
```


Our dataframe for clustering with include one row per player in our dataset and include 12 metrics.
```{r make df}
players <- unique(players_top$displayName)

res <- as.data.frame(matrix(NA, nrow = length(players), ncol = 10))

for(i in 1:length(players)){
  temp <- test[which(test$displayName == players[i]), ]
  
  res[i, 1:7] <- colMeans(temp[, clustering_columns], na.rm = TRUE) #average of all columns above
  res[i, 8] <- max(temp$max_acc, na.rm = TRUE) # max of max acceleration
  res[i, 9] <- max(temp$cum_load, na.rm = TRUE) # max of cumulative load
  res[i, 10] <- max(temp$avg_fatigueIndex, na.rm = TRUE) # max of average fatigue index
}

names(res) <- c("avg_max_acc", "avg_fatigueIndex", "avg_cum_load", "height", "weight", "age", "avg_load", "max_acc_max", "max_cum_load_in_game", "max_avg_fatigueIndex")

rownames(res) <- players
```

Remove NA values (before scaling).
All of our NA values are present in the 'age' column of our dataset. We will do so MICE imputation with the method Predictive Mean Matching ("PMM").
```{r}
sum(is.na(res))
sum(is.na(res$age))

library(mice)
imputed_data <- mice(res, m = 5, method = 'pmm', maxit = 50, seed = 500)
impute_data <- complete(imputed_data, action = 1)

sum(is.na(impute_data))
```

Scale clustering data.
```{r}
# avg_fatigueIndex is already in [0,1], so it does not need to be scaled.
unscaled_col <- impute_data$avg_fatigueIndex

all_data <- impute_data[, !names(impute_data) %in% c("avg_fatigueIndex")]

# Before scaling data, log transform.
all_data <- log(all_data + 1)

sdata <- scale(all_data)

sdata <- cbind(sdata, avg_fatigueIndex = unscaled_col)
```

Find the optimal number of clusters for analysis.
```{r}
library(cluster)
gapData <- clusGap(sdata, FUNcluster = kmeans, K.max = 15)

library(factoextra)
fviz_gap_stat(gapData)
```

Here, we see that the optimal number of clusters is only 2, but we will continue with three clusters to further see disparity between them.

Let's now run k-means clustering on the data.
```{r}
set.seed(12345) 

fit <- kmeans(x = sdata,
                centers = 3, 
                nstart = 25, 
                iter.max = 100)
clusters <- fit$cluster
centers <- fit$centers
summary(as.factor(clusters))
```

Let's move forward with three clusters.
```{r}
cluster <- c(1:3)
center_df <- data.frame(cluster, centers)

center_reshape <- gather(center_df, features, values, avg_max_acc:avg_fatigueIndex)
```


Heatmap
```{r}
heatmap <- ggplot(data = center_reshape, 
                   aes(x = features, y = cluster, fill = values)) + 
  scale_y_continuous(breaks = seq(1, 8, by = 1)) + 
  geom_tile() + 
  coord_equal() + 
  theme_bw() + 
  scale_fill_gradient2(low = "blue", 
                       mid = "white", 
                       high = "red",
                       midpoint =0, 
                       space = "Lab", 
                       na.value ="grey", 
                       guide = "colourbar", 
                       aesthetics = "fill") + 
  coord_flip() 

heatmap
```

Now, we can add total penalties to the clusters.

First, let's assign cluster number to player name.
```{r}
player_clusts <- data.frame(player_name = players, cluster = clusters)
rownames(player_clusts) <- NULL
```


```{r}
sum_pen <- as.data.frame(matrix(NA, nrow = nrow(player_clusts), ncol = 2))

for(i in 1:nrow(player_clusts)){
  temp <- test[which(test$displayName == player_clusts$player_name[i]), ]
  
  sum_pen[i, 1] <- sum(temp$tackle_pen, na.rm = TRUE)
  sum_pen[i, 2] <- sum(temp$formation_pen, na.rm = TRUE)
}

colnames(sum_pen) <- c("total_tackle_pen", "total_formation_pen")
sum_pen <- cbind(player_clusts, sum_pen)
```

Now that we have a sum of penalties for each player, we can count this total for each cluster and add it to our heatmap.
```{r}
totals <- as.data.frame(matrix(NA, nrow = 3, ncol = 5))

for(i in 1:3){
  temp <- sum_pen[which(sum_pen$cluster == i), ]
  
  totals[i, 1] <- sum(temp$total_tackle_pen, na.rm = TRUE)
  totals[i, 2] <- sum(temp$total_formation_pen, na.rm = TRUE)
  totals[i, 3] <- sum(nrow(temp))
  totals[i, 4] <- sum(temp$total_tackle_pen, na.rm = TRUE)/nrow(temp)
  totals[i, 5] <- sum(temp$total_formation_pen, na.rm = TRUE)/nrow(temp)
}

colnames(totals) <- c("total_tackle_pen", "total_formation_pen", "n_players", "tackle_pen_percent",
                      "formation_pen_percent")
```

add back to heatmap