---
title: "NFL_Fatigue_Analysis"
author: "Sarah Deussing"
date: "2024-11-11"
output: html_document
---

## Motivation
What is the problem you are considering and why is this an interesting problem? Has this problem been attempted before?

Our analysis will study fatigue in defensive linemen using their speed, acceleration, distance traveled, and orientation. We will be calculating a player load and analyze the defensive player's effectiveness. To do so, we will combine this movement data with play outcome data - if the player sacked the quarterback, had a solo tackle, etc.

The outcome of our analysis will be to predict likelihood of "mistakes" and positive plays, based on player fatigue and other game-time variables and to identify if specific players are more fatigue-resistant.

There has been published research on long-term fatigue and "burnout" in NFL players over the course of their careers, which focuses on long-term health and physiological effects. There has also been research on game day fatigue in European Football to determine limitations on performance based on fatigue, in order to create suitable, enhancing interventions (Alghannam). However, there is no player specific or NFL game day based fatigue analysis, which is what will set our project apart.


## Research

We wanted to find a measure of fatigue that was derived from tracking data. From an study on _Monitoring Fatigue During Intermittent Exercise With Accelerometer-Derived Metrics_ and Catapult's own description of Player Load, we found this idea of "Dynamic Load", that is a sum of triaxial acceleration across a set time period. We do not have acceleration in three plains but we do have an acceleration in the direction of play so we decided to derive a "Load per Play" value that will be acceleration summed per player per play.

As an extension to this, we will also derive a "Fatigue Index" based on the research by Beato et al., that will be a value of acceleration/speed for each value in time, i.


## Problem framing
What is the proposed solution to the problem?

Derive variables:

- Load by player by play
- Fatigue index by player by play

Perform a time-series analysis over the length of a game/period to see trend and seasonality of fatigue and the change in fatigue at specific events, e.g. "mistakes".

Define mistake as penalties that will be grouped based on type:
- Formation
  - Illegal Shift, Illegal Formation, Defensive Offside

- Tackle
  - Defensive Holding, Horse Collar Tackle, Low Block, Illegal Use of Hands, Roughing the Passer

Use player demographic data, time, tracking and load data to predict probability of "mistake" at a given time.

Use player demographic data, time, tracking and load data to predict positive defensive outcomes such as sacks, pressure and fumbles.

Identify fatigue-resistant players and cluster to see if they have similar demographic characteristics.

## Data overview
What is the dataset you are going to use to solve this problem? Describe the dataset and its characteristics.

The dataset is a part of the NFL Big Data Bowl 2025, and it is taken from:
https://www.kaggle.com/competitions/nfl-big-data-bowl-2025/data 

We are going to use three main parts of the dataset: player_play, players, and the tracking data for each week.

The player_play data contains information about the action of the play, including its outcome or any penalties. Each row corresponds to one player's individual actions for each play within a game.

The plays data contains information about each player, including position, height, weight, and college. We will use this data to filter to only defensive linemen within all data sets.

The tracking data has the x-coordinate, y-coordinate, direction, orientation, speed, and acceleration for each player involved in each play in a game. We will use this movement data to calculate player fatigue. This tracking data is available for all 9 weeks of the season.

## Contribution
Imogen Meers & Sarah Deussing

## Bibliography

Alghannam, Abdullah F. “Metabolic limitations of performance and fatigue in football.” Asian journal of sports medicine vol. 3,2 (2012): 65-73. doi:10.5812/asjsm.34699

Beato, Marco et al. “Monitoring Fatigue During Intermittent Exercise With Accelerometer-Derived Metrics.” Frontiers in physiology vol. 10 780. 26 Jun. 2019, doi:10.3389/fphys.2019.00780

Finster-Rowen, Alex. "What is Player Load?" Catapult Sports. Catapult Sports, n.d. Web.


## Implementation
```{r}
library(dplyr)
```


#### Data
Play Data for Each Player
```{r}
play <- read.csv('player_play.csv')
player_data <- read.csv('players.csv')
```


Filter to only defensive linemen.
- DE (Defensive End)
- DT (Defensive Tackle)
- NT (Nose Tackle)
```{r}
def_line <- player_data[player_data$position == 'DE' | player_data$position == 'DT' |
                          player_data$position == 'NT',]

def_line <- play %>%
  left_join(def_line[, c("nflId", "position")], by = "nflId")

def_line <- def_line %>%
  select(-c('hadRushAttempt', 'rushingYards', 'hadDropback', 'passingYards', 'sackYardsAsOffense',
            'hadPassReception', 'receivingYards', 'wasTargettedReceiver', 'yardageGainedAfterTheCatch',
            'fumbles', 'fumbleLost', 'fumbleOutOfBounds', 'wasInitialPassRusher', 'wasRunningRoute',
            'routeRan', 'blockedPlayerNFLId1', 'blockedPlayerNFLId2', 'blockedPlayerNFLId3'))
```

Variables for defensive linemen: 
 - assistedTackle: required an assist to make a tackle
 - forcedFumbleAsDefense: forced a fumble by the opposing team
 - halfSackYardsAsDefense: yards conceded by the offense because of a half-sack
 - passDefensed: a passing play was stopped by the player
 - quarterbackHit: player recorded a QB hit
 - sackYardsAsDefense: yards conceded by the offense because of a sack by the player
 - safetyAsDefense: player forced a safety on this play
 - soloTackle: player recorded a solo tackle on this play
 - tackleAssist: player was awarded an assisted tackle
 - tackleForALoss: player recorded a tackle behind the line of scrimmage
 - tackleForALossYardage: yards conceded by the offense because of a tackle
 - hadInterception: player intercepted a pass
 - interceptionYards: yards returned by the player on an intercepted pass
 - fumbleRecoveries: number of fumbles recovered by the player
 - fumbleRecoveryYards: yards returned by the player on a fumble recovery
 - penaltyNames
 - causedPressure: player pressured the QB
 - timeToPressureAsPassRusher: time elapsed from snap to player reaching a pressure pro. >= to 0.75
 - inMotionAtBallSnap: player was in motion at snap
 - shiftSinceLineset: player shifted since the lineset
 - pressureAllowedAsBlocker: any pass rushers that the blocker had a true matchup against recorded a pressure
 - timeToPressureAllowedAsBlocker: time from snap to pass rusher who blocker had true matchup against achieving a pressure prob. > 0.75
 - pff_defensiveCoverageAssignment: defensive coverage assignment given to the player
 
```{r}
summary(def_line)
```


Tracking
```{r, eval = FALSE}
week1 <- read.csv('tracking_week_1.csv')
```
 
Fatigue variables:
 - s (speed), a (acceleration), dis (distance traveled), dir (direction), o (orientation) 
 
Join Data
```{r}
data <- inner_join(def_line, week1, by = c("gameId", "playId", "nflId"))
colnames(data)
```

Player Load = Acceleration per linemen (per play)
```{r}
data <- data %>%
  group_by(gameId, playId, nflId) %>%
  mutate(load = sum(a, na.rm = TRUE)) %>% ungroup()
```

Plot load Per Player for a game
```{r}
game <- data[data$gameId == 2022090800,]

for (id in unique(game$nflId)) {
  player_data <- game %>% filter(nflId == id)
  p <- ggplot(player_data, aes(x = playId, y = load)) +
    geom_point() + 
    geom_smooth(method = "loess", color = "blue") + 
    labs(x = "Play ID", y = "Load", title = paste0("Player Load in Game - NFL ID: ", id)) +
    theme_minimal()

  print(p)
}
```


```{r Tackles over Time}
library(lubridate)
library(tidyr)
library(ggplot2)
head(data)

tackle_pen <- c("Horse Collar Tackle", "Defensive Holding", "Low Block", "Illegal Use of Hands", "Roughing the Passer")
formation_pen <-  c("Illegal Shift", "Illegal Formation", "Defensive Offside")

play_details <- read.csv("plays.csv") %>% select(gameId, playId, quarter, gameClock)
head(play_details)

# penalties <- play %>% select(gameId, playId, nflId, penaltyNames) %>% full_join(play_details) %>% mutate(penaltyType = ifelse(penaltyNames %in% tackle_pen, "Tackle", ifelse(penaltyNames %in% formation_pen, "Formation", NA)), totalClock = (quarter-1) *  ms("15:00") +(ms("15:00")- ms(gameClock))) 


penalties <- play %>% select(gameId, playId, nflId, penaltyNames) %>% full_join(play_details) %>% mutate(penaltyType = ifelse(penaltyNames %in% tackle_pen, "Tackle", ifelse(penaltyNames %in% formation_pen, "Formation", NA)))
                                                                                                         
plot_data <- penalties %>% select(quarter, penaltyType) %>% na.omit() %>% group_by(quarter, penaltyType) %>% summarise(count = n()) %>% ungroup() 

stacked_bar <- ggplot(plot_data, aes(x = as.factor(quarter), y = count, fill = penaltyType)) +
  geom_bar(stat = "identity") +
  labs(title = "Penalties by Category and Quarter", x = "Quarter", y = "Count", fill = "Penalty Type") +
  theme_minimal()

stacked_bar

# plot_data <- plot_data %>% na.omit()
# plot_data <- plot_data %>% pivot_wider(names_from = penaltyType, values_from = count, values_fill = list(count = 0))
# 
# plot_data <- penalties %>% select(gameId, playId, penaltyType) %>% distinct() %>% arrange(gameId)
# unique(plot_data$playId)
# 
# ggplot(data = plot_data) + geom_density(aes(x = totalClock, y = count, color = penaltyType))
# tail(penalties)
# unique(penalties$totalClock)
# ms("15:00")- ms(penalties$gameClock)

```
This doesn't show exactly what we thought. There is an increase in both types of penalties from the first to the second quarter, and an increase in formation penalties from the third to fourth quarter. We would expect some type of reset after half time, so we will also look at fatigue by quarter. This is just a plot for week 1 data so it's not totally representative of the season or the "NFL" in general.
